### 2025-05-07 paper/paper.tex
Update paper.tex

\section{Introduction}
This paper will present the relevant literature in the process of traffic signal control using machine learning techniques, along with the timeline of research developments over time.

@@

Early breakthroughs in large-scale MARL focused on a more difficult problem: urban traffic management. This initially came with a new open-sourced model, CityFlow, a multi-agent reinforcement model which could handle large road networks with thousands of intersections in a simulated environment \cite{7}. These large-scale models have shown significant improvements in performance while working at a larger scale, making them more adaptable to real situations.

@@

A third research gap appears to be the influence of various real-world distractions that can occur, ranging from road accidents, pedestrians, and emergency vehicles. Although Du et al. have begun to take into account pedestrians \cite{13}, there are still several factors that remain unaccounted for in the traffic signal prediction process, which can back up traffic worse than threshold options.

@@

Each intersection is managed by an instance of {\tt DQNAgent} (in policies.py), which maintains both an \emph{online} network and a \emph{target} network. At initialization, we load the target’s weights from the online network and set up an Adam optimizer (learning rate 0.001), a discount factor $\gamma=0.99$, and an $\varepsilon$‐greedy exploration schedule that decays $\varepsilon$ from 1.0 to 0.05 by a factor of 0.995 per learning step. Experiences are stored in a fixed‐capacity deque buffer (size 5,000), and we sample mini‐batches of 32 transitions once the buffer is sufficiently large. In each learn() call we compute the predicted Q‐values for state–action pairs, form TD‐targets via 
\[
Q_{\mathrm{target}} \;=\; r \;+\; \gamma\,(1 - d)\,\max_{a'} Q_{\mathrm{target}}(s', a')
\]
and minimize the mean‐squared error between predictions and targets. This target arises directly from the Bellman optimality equation and serves as the reference value toward which our online Q‐network is trained. Below is a concise, technical explanation of each component. The scalar \textit{r} represents the instantaneous reward signal returned by the environment upon executing action \textit{a} in state \textit{s}. In traffic‐signal control, we typically define \textit{r} as a negative congestion metric (e.g., the sum of queue lengths), so that higher rewards correspond to smoother traffic flow. This term anchors the TD target in the actual, observed outcome of the agent’s decision. $\gamma \in [0,1)$ is the temporal‐discount coefficient that geometrically attenuates the contribution of future returns.  
A value of $\gamma = 0.99$ indicates that the agent values rewards received one step later at $99\%$ of their original magnitude.  
By tuning $\gamma$, one controls the trade‐off between short‐term gains and long‐horizon optimization. Lower $\gamma$ leads to myopic behavior, while values approaching 1 emphasize long‐term planning. The binary flag $d$ equals $1$ if the sampled transition terminates an episode (i.e., $s'$ is terminal), and $0$ otherwise.  
Multiplying by $(1 - d)$ ensures that, in terminal states, no bootstrapped future value is added, thus strictly enforcing that $Q_{\mathrm{target}} = r$ when the episode ends. This prevents spurious propagation of post‐terminal $Q$-values. The operator $\max_{a'} Q(s',a')$ extracts the highest $Q$‐value estimate across all possible actions $a'$ in the successor state $s'$, as evaluated by the target network. This “greedy” selection embodies the assumption that, after the current transition, the agent will henceforth behave optimally, consistently choosing the action with maximal expected return. By summing the immediate reward $r$ and the discounted maximal next‐state value (when $d = 0$), we obtain a single‐step bootstrap estimate of the true return. During training, the online network’s prediction $Q_{\mathrm{pred}}(s,a)$ is regressed toward $Q_{\mathrm{target}}$ via mean‐squared error. Periodic synchronization of the target network stabilizes training by holding the bootstrap reference fixed for a configurable number of gradient updates. After each gradient update, we decay $\varepsilon$, and every 100 updates we synchronize the target network with the online network. To scale this to multiple intersections, we wrap several {\tt DQNAgent} objects in {\tt MultiDQNAgent}, forwarding action selection, memory storage, and learning calls independently to each agent.

@@

\subsection{Cross Validation}

One potential concern over the implementation of our model is not the model itself, but the trustworthiness of the results. As a result, the design of the model took careful consideration into preparing valid results through cross validation and critical analysis. Various situations were considered for training, while only a single traffic scenario was used for testing and benchmarking.

The first way we validated our results is by creating new scenarios. Each Episode of the model sees a completely new traffic scenario within Starkville, MS, granted with similar size due to computational limitations of large scale traffic. We then printed the reward, the punishment part of the Q-score each episode the ensure progress and exploration, and the results reveal relative consistency with a strong balance of exploration and exploitation. Even with new situations, the score grew and grew over time (please note that the reward is negative). The model was trained for 150 episodes in this scenario, to reach convergence while exploring as much as possible.

@@

During our benchmarking phase we generated one testing scenario. This scenario uses a moderate-light traffic, based on Starkville, Mississippi's population and is used as a baseline to ensure that the improved results are not a result of a more or less congested traffic pattern. When we run the benchmarks we carefully ensured that this was performed to make sure results are trustworthy, and the the model did not optimize only one single situation it saw. When doing this, the results were overwhelmingly positive, and ensures that MARL based Deep Q-Learning is a valid way to optimize traffic signal controllers, as seen in section ``Results over baseline.``

@@

The MARL Deep Q-learning model saw notable improvements in terms of responsiveness. When there were no signs of incoming traffic or lines in stopped lanes, it skipped the redundancy of changing the state of traffic lights resulting in a generally smoother flow of traffic. When it did change, it seemed intelligent, considering how long cars had been waiting along with how many cars were waiting. While our model did not allow collisions to be considered, the more continuous flow of traffic saw fewer emergency stops, and generally safer roads. Analysis also revealed that vehicles would often arrive at their destination sooner than their threshold counterparts, with less time spent on the road generally resulting in safer roads with less congestion.

@@

Testing the average trip time was one of the most important benchmarks implemented, since longer trips result in less cars on the road and a shorter time spent on the road, the odds of a traffic incident decreases dramatically. When using traditional threshold signals the average trip time took 695.34 seconds, or 11 minutes and 34 seconds. When using neural network based traffic signals a notable decrease in average trip time was observed, decreasing to 587.61 seconds, or 9 minutes and 47 seconds. With a nearly 108 second decrease in average trip time, we saw a 15\% reduction in average trip time.

We also analyzed time spent waiting for both benchmarks, to see where the decrease in average trip time came from. In theory, this result would come from a decreased wait time, but some of it could come from the decreased traffic and increased speeds associated with better traffic signal management. With threshold signals, the average time spent waiting was 298.4 seconds, but with neural networks this decreased to 195.76 seconds. This resulted in a decrease of 102 seconds, confirming that the decreased wait time significantly contributed to the lower average trip time.

Another important metric to decrease it the average length of stopped cars at any point in time, known as the average queue length. Because our wait time decreased, we can expect a notable decrease in average queue length, as less cars need to stop and wait. This is confirmed, as we achieve an average queue length of 230 vehicles with threshold neural networks, and a queue length of 55 vehicles when using neural networks. This is an astronomical decrease, with an average of 175 less vehicles being stopped at any point in time.  This is a nearly 76\% decrease in average queue length, indicating a significant improvement in performance over baselines.

Finally, the maximum length of cars was determined in order to ensure that traffic did not ever back up very bad. Because the average queue length decreased greatly, it is safe to assume that the maximum queue length decreased significantly, which is confirmed by our results. When using traditional threshold signals the maximum queue length reached 1373 vehicles, and the maximum queue length reached 316 vehicles. This indicates that the peak congestion was significantly improved through our model over baselines, preventing large backups which can be frustrating and potentially dangerous.

@@

With the average time spent driving decreasing drivers are generally safer, and large sections of traffic are less likely to make sudden stops. The shorter queue lengths result in less frustrated drivers, and quicker driving times, which also indicates a smoother flow of traffic and safer roads. This did come at the cost of computational power, resulting in a slower model that took time to train. It does perform faster than real time, but given the wrong situation could slow down the model where it could not make real time predictions; however, because the results were overall positive, there likely exists a cheaper and more efficient way to direct traffic.

While deep Q-Learning with MARL objectively performed well, computation times were relatively slow and implementation would prove very costly. Because of this, the MDOT should also investigate sensor based traffic signal controllers. These signals are cheaper to install, but consider timers and congestion levels with more strictly defined rules. Implementing these would avoid the high cost of training, implementation, and upkeep. Because using state data increased the performance with neural networks, its likely that a simple understanding of the state of the road could help direct traffic without implementing an expensive and unpredictable model. While still having some cost, it would decrease time spent driving and keep the roads of Mississippi safer.

@@

In order to download the code one would need to go to GitHub, and view chandlerweeks/TrafficSignalManagement. The link to the repository is: https://github.com/ChandlerWeeks/TrafficSignalManagement. Once here, fork the repository and make a local clone on your device. If a release is available at the time of reading the paper, that link will allow direct access of the codebase.

@@

Once every prerequisite has been setup, to test installation and configuration, one should open SUMO-GUI and open the file in data/simulation.sumocfg. If the file runs and traffic is observed, then the testing traffic flow scenario has been loaded and the code is ready to run.

@@

In order to run the benchmarks with the model, one must run the trained model first to get the trip information. This can be ran through this command \texttt{python -m traffic\_signal\_nn.eval.run\_model --config traffic\_signal\_nn/config/config\_dqn\_city.ini}, which runs the trained model on the same situation as the previous benchmark, for consistency as mentioned in cross-validation. Just like without the neural network, the queue length is not obtained for the benchmarks, so one will need to run \texttt{python .\textbackslash traffic\_signal\_nn\textbackslash eval\textbackslash get\_model\_queue\_length.py --config .\textbackslash traffic\_signal\_nn\textbackslash config\textbackslash config\_dqn\_city.ini --output .\textbackslash data\textbackslash output\textbackslash queue\_length.csv} in order the get the updated queue lengths. Once all of this has been done, one can simply run \texttt{python benchmark.py} to get the benchmark results for the trained model.

Setup of the project can be rather time consuming, as mentioned, a video can provided to the reviewers running the script completely setup if necessary.

@@

Large language models were used in assistance of coding, particularly for debugging and assistance with small code snippets, but never used to generate large parts of a function, or codebits. The code is developed by human, but AI assisted with some lines.

>>>>

\section{Introduction}
This paper will present the relevant literature on the process of traffic signal control using machine learning techniques, along with the timeline of research developments over time.

@@

Early breakthroughs in large-scale MARL focused on a more difficult problem: urban traffic management. This initially came with a new open-sourced model, CityFlow, a multi-agent reinforcement model that could handle large road networks with thousands of intersections in a simulated environment \cite{7}. These large-scale models have shown significant improvements in performance while working at a larger scale, making them more adaptable to real situations.

@@

A third research gap appears to be the influence of various real-world distractions that can occur, ranging from road accidents, to pedestrians, and emergency vehicles. Although Du et al. have begun to take into account pedestrians \cite{13}, there are still several factors that remain unaccounted for in the traffic signal prediction process, which can back up traffic worse than threshold options.

@@

Each intersection is managed by an instance of {\tt DQNAgent} (in policies.py), which maintains both an \emph{online} network and a \emph{target} network. At initialization, we load the target’s weights from the online network and set up an Adam optimizer (learning rate 0.001), a discount factor $\gamma=0.99$, and an $\varepsilon$‐greedy exploration schedule that decays $\varepsilon$ from 1.0 to 0.05 by a factor of 0.995 per learning step. Experiences are stored in a fixed‐capacity deque buffer (size 5,000), and we sample mini‐batches of 32 transitions once the buffer is sufficiently large. In each learn() call we compute the predicted Q‐values for state–action pairs, form TD‐targets via 
\[
Q_{\mathrm{target}} \;=\; r \;+\; \gamma\,(1 - d)\,\max_{a'} Q_{\mathrm{target}}(s', a')
\]
and minimize the mean‐squared error between predictions and targets. This target arises directly from the Bellman optimality equation and serves as the reference value toward which our online Q‐network is trained. Below is a concise, technical explanation of each component. The scalar \textit{r} represents the instantaneous reward signal returned by the environment upon executing action \textit{a} in state \textit{s}. In traffic‐signal control, we typically define \textit{r} as a negative congestion metric (e.g., the sum of queue lengths), so that higher rewards correspond to smoother traffic flow. This term anchors the TD target in the actual, observed outcome of the agent’s decision. $\gamma \in [0,1)$ is the temporal‐discount coefficient that geometrically attenuates the contribution of future returns.  
A value of $\gamma = 0.99$ indicates that the agent values rewards received one step later at $99\%$ of their original magnitude.  
By tuning $\gamma$, one controls the trade‐off between short‐term gains and long‐horizon optimization. Lower $\gamma$ leads to myopic behavior, while values approaching 1 emphasize long‐term planning. The binary flag $d$ equals $1$ if the sampled transition terminates an episode (i.e., $s'$ is terminal), and $0$ otherwise.  
Multiplying by $(1 - d)$ ensures that, in terminal states, no bootstrapped future value is added, thus strictly enforcing that $Q_{\mathrm{target}} = r$ when the episode ends. This prevents spurious propagation of post‐terminal $Q$-values. The operator $\max_{a'} Q(s',a')$ extracts the highest $Q$‐value estimate across all possible actions $a'$ in the successor state $s'$, as evaluated by the target network. This “greedy” selection embodies the assumption that, after the current transition, the agent will henceforth behave optimally, consistently choosing the action with the maximal expected return. By summing the immediate reward $r$ and the discounted maximal next‐state value (when $d = 0$), we obtain a single‐step bootstrap estimate of the true return. During training, the online network’s prediction $Q_{\mathrm{pred}}(s,a)$ is regressed toward $Q_{\mathrm{target}}$ via mean‐squared error. Periodic synchronization of the target network stabilizes training by holding the bootstrap reference fixed for a configurable number of gradient updates. After each gradient update, we decay $\varepsilon$, and every 100 updates we synchronize the target network with the online network. To scale this to multiple intersections, we wrap several {\tt DQNAgent} objects in {\tt MultiDQNAgent}, forwarding action selection, memory storage, and learning calls independently to each agent.

@@

\subsection{Cross-Validation}

One potential concern over the implementation of our model is not the model itself, but the trustworthiness of the results. As a result, the design of the model took careful consideration into preparing valid results through cross-validation and critical analysis. Various situations were considered for training, while only a single traffic scenario was used for testing and benchmarking.

The first way we validated our results is by creating new scenarios. Each episode of the model sees a completely new traffic scenario within Starkville, MS, granted of a similar size due to computational limitations of large-scale traffic simulation. We then printed the reward, and the punishment part of the Q-score for each episode to ensure progress and exploration, and the results reveal relative consistency with a strong balance of exploration and exploitation. Even with new situations, the score grew and grew over time (please note that the reward is negative). The model was trained for 150 episodes in this scenario, to reach convergence while exploring as much as possible.

@@

During our benchmarking phase, we generated one testing scenario. This scenario uses moderate-light traffic, based on Starkville, Mississippi's population, and is used as a baseline to ensure that the improved results are not a result of a more or less congested traffic pattern. When we ran the benchmarks we carefully ensured that this was performed to make sure results were trustworthy, and the model did not optimize only one single situation it saw. When doing this, the results were overwhelmingly positive, and ensures that MARL-based Deep Q-Learning is a valid way to optimize traffic signal controllers, as seen in the section ``Results over baseline.``

@@

The MARL Deep Q-learning model saw notable improvements in terms of responsiveness. When there were no signs of incoming traffic or lines in stopped lanes, it skipped the redundancy of changing the state of traffic lights resulting in a generally smoother flow of traffic. When it did change, it seemed intelligent, considering how long cars had been waiting along with how many cars were waiting. While our model did not allow collisions to be considered, the more continuous flow of traffic saw fewer emergency stops, and generally safer roads. The analysis also revealed that vehicles would often arrive at their destination sooner than their threshold counterparts, with less time spent on the road generally resulting in safer roads with less congestion.

@@

Testing the average trip time was one of the most important benchmarks implemented, since longer trips resulted in fewer cars on the road and a shorter time spent on the road, the odds of a traffic incident decreased dramatically. When using traditional threshold signals the average trip time took 695.34 seconds, or 11 minutes and 34 seconds. When using neural network-based traffic signals a notable decrease in average trip time was observed, decreasing to 587.61 seconds, or 9 minutes and 47 seconds. With a nearly 108-second decrease in average trip time, we saw a 15\% reduction in average trip time.

We also analyzed time spent waiting for both benchmarks, to see where the decrease in average trip time came from. In theory, this result would come from a decreased wait time, but some of it could come from the decreased traffic and increased speeds associated with better traffic signal management. With threshold signals, the average time spent waiting was 298.4 seconds, but with neural networks, this decreased to 195.76 seconds. This resulted in a decrease of 102 seconds, confirming that the decreased wait time significantly contributed to the lower average trip time.

Another important metric to decrease is the average length of stopped cars at any point in time, known as the average queue length. Because our wait time decreased, we can expect a notable decrease in average queue length, as fewer cars need to stop and wait. This is confirmed, as we achieve an average queue length of 230 vehicles with threshold neural networks, and a queue length of 55 vehicles when using neural networks. This is an astronomical decrease, with an average of 175 fewer vehicles being stopped at any point in time.  This is a nearly 76\% decrease in average queue length, indicating a significant improvement in performance over baselines.

Finally, the maximum length of cars was determined in order to ensure that traffic did not ever back up very badly. Because the average queue length decreased greatly, it is safe to assume that the maximum queue length decreased significantly, which is confirmed by our results. When using traditional threshold signals the maximum queue length reached 1373 vehicles, and the maximum queue length reached 316 vehicles. This indicates that the peak congestion was significantly improved through our model over baselines, preventing large backups which can be frustrating and potentially dangerous.

@@

With the average time spent driving decreasing drivers are generally safer, and large sections of traffic are less likely to make sudden stops. The shorter queue lengths result in less frustrated drivers, and quicker driving times, which also indicates a smoother flow of traffic and safer roads. This did come at the cost of computational power, resulting in a slower model that took time to train. It does perform faster than real-time, but given the wrong situation, could slow down the model to where it could not make real-time predictions. However, because the results were overall positive, there likely exists a cheaper and more efficient way to direct traffic.

While deep Q-Learning with MARL objectively performed well, computation times were relatively slow and implementation would prove very costly. Because of this, the MDOT should also investigate sensor-based traffic signal controllers. These signals are cheaper to install, but consider timers and congestion levels with more strictly defined rules. Implementing these would avoid the high cost of training, implementation, and upkeep. Because using state data increased the performance with neural networks, it's likely that a simple understanding of the state of the road could help direct traffic without implementing an expensive and unpredictable model. While still having some cost, it would decrease time spent driving and keep the roads of Mississippi safer.

@@

In order to download the code one would need to go to GitHub and view chandlerweeks/TrafficSignalManagement. The link to the repository is: https://github.com/ChandlerWeeks/TrafficSignalManagement. Once here, fork the repository and make a local clone on your device. If a release is available at the time of reading the paper, that link will allow direct access to the codebase.

@@

Once every prerequisite has been set up, to test installation and configuration, one should open SUMO-GUI and open the file in data/simulation.sumocfg. If the file runs and traffic is observed, then the testing traffic flow scenario has been loaded and the code is ready to run.

@@

In order to run the benchmarks with the model, one must run the trained model first to get the trip information. This can be run through this command \texttt{python -m traffic\_signal\_nn.eval.run\_model --config traffic\_signal\_nn/config/config\_dqn\_city.ini}, which runs the trained model on the same situation as the previous benchmark, for consistency as mentioned in cross-validation. Just like without the neural network, the queue length is not obtained for the benchmarks, so one will need to run \texttt{python .\textbackslash traffic\_signal\_nn\textbackslash eval\textbackslash get\_model\_queue\_length.py --config .\textbackslash traffic\_signal\_nn\textbackslash config\textbackslash config\_dqn\_city.ini --output .\textbackslash data\textbackslash output\textbackslash queue\_length.csv} in order the get the updated queue lengths. Once all of this has been done, one can simply run \texttt{python benchmark.py} to get the benchmark results for the trained model.

Setup of the project can be rather time-consuming, as mentioned, a video can provided to the reviewers running the script completely setup if necessary.

@@

Large language models were used in assistance of coding, particularly for debugging and assistance with small code snippets, but never used to generate large parts of a function or code bits. The code is developed by humans, but AI assisted with some lines.
